---
title: "ABCs data import"
author: "Dan Weinberger"
date: '2022-09-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(plotly)
#install.packages("waffle", repos = "https://cinc.rud.is") 
library(waffle)
library(reshape2)
source('./R/process_ecdc.R')
```


```{r}
#a1 <- read.csv('https://data.cdc.gov/api/views/en3s-hzsr/rows.csv?accessType=DOWNLOAD')

#saveRDS(a1,'./Data/ABCs_data_summary_2024_01_01.rds')

a1 <-readRDS('./Data/ABCs_data_summary_2024_01_01.rds')
```

Based on ABCs Bact-Facts, projected 3250 IPD deaths nationally in 2019. In 2019 there were 810 deaths nationally recorded as J13 (pneumococcal pneumonia) or A40.3 (pneumococcal septicemia) or G00.1 (pneumococcal meningitis) from CDC WONDER (B95.3 was not on WONDER). With the line list data we get 923 with J13, A40.3, G00.1, B95.3. There were 537 Streptococcal sepsis (A40.8/A40.9)


ABCS genomic data are in: https://www.ncbi.nlm.nih.gov/bioproject/284954

```{r}
p1 <- a1 %>%
  filter(Topic =='Case Rates' & ViewBy=='Age' & Year>=1998) %>%
 ggplot( aes(x=Year, y=Value, col=ViewBy2)) +
  geom_line() +
  theme_classic() 

ggplotly(p1)
```

```{r}
p2 <- a1 %>%
  filter(Topic =='Death Rates' & ViewBy=='Age' & Year>=1998) %>%
 ggplot( aes(x=Year, y=Value, col=ViewBy2)) +
  geom_line() +
  theme_classic() 

ggplotly(p2)
```


```{r, fig.width=12, fig.height=4}
b2019 <- read.csv('./Data/ABCs/jiac058_suppl_supplementary_table_s2.csv') %>%
  group_by(State, sero, LT5) %>%
  summarize(N_cases=n()) %>%
  mutate(sero=as.factor(sero)) %>%
  ungroup() %>%
  group_by(State, LT5) %>%
  mutate(pct = N_cases/sum(N_cases)) %>%
  ungroup()
  
ggplotly(ggplot(b2019[b2019$LT5=='neg',], aes(fill=State, y=pct, x=sero)) + 
    geom_bar(position="dodge", stat="identity") 
) 
```
correlations
```{r}

adult.c <- b2019 %>%
  filter(LT5=='neg') %>%
  ungroup() %>%
  group_by(sero) %>%
  mutate(tot_N_cases=sum(N_cases)) %>%
  ungroup() %>%
  filter(tot_N_cases >= 10) %>%
 dcast(  State ~ sero, value.var='pct') %>%
  replace(is.na(.), 0)

cor(adult.c[,-1])
```



```{r}


  p3 <- ggplot(b2019[b2019$LT5=='neg',], aes(fill = sero, values = N_cases)) +
    geom_waffle(color='white', size = 0.25) +
  facet_wrap(~State) +
  theme_classic()
ggplotly(p3)
```


```{r}
ecdc <- process_ecdc() %>%
  filter(agegrp==5 & country != 'EU/EEA' & total_cases >100 & !(st %in% c('11','11F','19','36','39','43','7','17','Other')))

ecdc.c <-  dcast(ecdc  ,country ~ st, value.var='pct') %>%
  replace(is.na(.), 0)

ecdc.c.scale <- apply(ecdc.c[,-1],2,scale)
corr.ecdc <- cor(ecdc.c.scale)

dist_mat <- dist(t(ecdc.c.scale), method = 'euclidean')
hclust_avg <- hclust(dist_mat, method = 'average')
plot(hclust_avg)
View(corr.ecdc)
```

Principal Components
```{r}

a1 <- prcomp(ecdc.c[,-1], scale = TRUE)

#reverse the signs
a1$rotation <- -1*a1$rotation

#display principal components
a1$rotation

biplot(a1, scale = 0, cex=0.5)

#calculate total variance explained by each principal component
round(a1$sdev^2 / sum(a1$sdev^2)*100)

```


## US ABCs serotype data 2016-21
https://data.cdc.gov/Public-Health-Surveillance/2016-2021-Serotype-Data-for-Invasive-Pneumococcal-/qvzb-qs6p/data_preview
```{r}
st1 <- read.csv('C:/Users/DMW63/Desktop/R_Projects/ABCs_pneumococcal_data/Data/2016-2021_Serotype_Data_for_Invasive_Pneumococcal_Disease_Cases_by_Age_Group_from_Active_Bacterial_Core_surveillance_20240101.csv') 

```

```{r}
ecdc2 <- process_ecdc2() %>%
  filter( country != 'EU/EEA' & total_cases >100 & !(st %in% c('11','11F','19','36','39','43','7','17','Other')))

uk_data <- ecdc2 %>%
  filter(country=='United Kingdom'  & year %in% c(2011,2019) & st !='MISS') %>%
  group_by(agegrp,st,year) %>%
  summarize(N_IPD=sum(N_cases)) %>%
  ungroup() %>%
  group_by(agegrp,year) %>%
  arrange(-N_IPD) %>%
  mutate(stN=row_number(),
        pcv20 = if_else(st %in% c('1', '3', '4', '5', '6A', '6B', '7F', '8', '9V', '10A', '11A', '12F', '14', '15B', '18C', '19A', '19F', '22F', '23F','33F'),'PCV20','Other')) %>%
  ungroup()

```

```{r, fig.width=5, fig.height=3}
 p1 <- uk_data %>% 
   filter(agegrp==1 & year==2011) %>%
   ggplot(aes(x=stN, y=N_IPD, color=pcv20)) +
  #geom_line() +
  geom_text(aes(label=st), size=2, angle=45)+
  theme_minimal()+
  ylab('Number of isolates')+
  xlab('Serotype order')+
  ggtitle('IPD, children <5 in the UK, 2011')
p1

 p2 <- uk_data %>% 
   filter(agegrp==1 & year==2019) %>%
   ggplot(aes(x=stN, y=N_IPD, color=pcv20)) +
  #geom_line() +
  geom_text(aes(label=st), size=2, angle=45)+
  theme_minimal()+
  ylab('Number of isolates')+
  xlab('Serotype order')+
  ggtitle('IPD, children <5 in the UK, 2019')
p2

```

