---
title: "ABCs data import"
author: "Dan Weinberger"
date: '2022-09-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(viridis)
library(plotly)
library(scales)
library(pheatmap)
#install.packages("waffle", repos = "https://cinc.rud.is") 
library(waffle)
library(reshape2)
source('./R/process_ecdc.R')
```



##ABCs serotype specific IPD data by age!
```{r, fig.width=5, fig.height=10}
# s1 <- read.csv('https://data.cdc.gov/api/views/qvzb-qs6p/rows.csv?accessType=DOWNLOAD')
# saveRDS(s1,'./Data/ABCs_st_1998_2021.rds') 

s1 <- readRDS('./Data/ABCs_st_1998_2021.rds') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  mutate( st= if_else(st=='16','16F', st)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup()


p1 <- s1 %>%
    group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ggplot(aes(x=year, y=log(N_IPD+0.5), group=st, color=st)) +
  geom_line()+
  theme_minimal()
ggplotly(p1)

s2_pre<- s1 %>%
  filter(year %in% c(1998, 1999)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup() %>%
    group_by(st) %>%
  summarize(N_IPD_pre=mean(N_IPD)) %>%
  ungroup() %>%
    tidyr::complete(st,  fill=list(N_IPD_pre=0))  #fills 0

s2_pre13<- s1 %>%
  filter(year %in% c(2008, 2009)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup() %>%
    group_by(st) %>%
  summarize(N_IPD_pre=mean(N_IPD)) %>%
  ungroup() %>%
    tidyr::complete(st,  fill=list(N_IPD_pre=0))  #fills 0

s2<- s1 %>%
    group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>% #sum across age group
  ungroup() %>%
    tidyr::complete(st, year, fill=list(N_IPD=0))%>%  #fills 0
  left_join(s2_pre, by='st') %>%
  mutate(N_IPD_pre = if_else(is.na(N_IPD_pre),0, N_IPD_pre) ,
    logRR = log((N_IPD+1)/(N_IPD_pre+1) )) 

max_RR <- s2 %>%
  group_by(st) %>%
  summarize(max_RR = max(logRR)) %>%
  arrange((max_RR))

s2$st <- factor(s2$st, levels = max_RR$st)

ggplot(s2, aes(x = factor(year), y = st, fill = logRR)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(x = "Year", y =
         
         "serotype", fill = "N_IPD") +
  theme_minimal()


df_wide <- s2 %>%
  dplyr::select(year, st, logRR) %>%
  tidyr::pivot_wider(names_from = year, values_from = logRR)

pheatmap(
  df_wide[,-1],
  color = viridis(100),
  cluster_rows = TRUE,
  cluster_cols = F,
  scale = "none",
  border_color = NA,
  labels_row=df_wide$st
)


```

pre-PCV13 heatmap only
```{r, fig.width=5, fig.height=10}
df_wide <- s2 %>%
  filter(year<=2009) %>%
  dplyr::select(year, st, logRR) %>%
  tidyr::pivot_wider(names_from = year, values_from = logRR)

pheatmap(
  df_wide[,-1],
  color = viridis(100),
  cluster_rows = TRUE,
  cluster_cols = F,
  scale = "none",
  border_color = NA,
  labels_row=df_wide$st
)
```

##post PV13 only

```{r, fig.width=5, fig.height=10}
s2_13<- s1 %>%
    group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>% #sum across age group
  ungroup() %>%
    tidyr::complete(st, year, fill=list(N_IPD=0))%>%  #fills 0
  left_join(s2_pre13, by='st') %>%
  mutate(N_IPD_pre = if_else(is.na(N_IPD_pre),0, N_IPD_pre) ,
    logRR13 = log((N_IPD+1)/(N_IPD_pre+1) )) 


df_wide13 <- s2_13 %>%
  dplyr::select(year, st, logRR13) %>%
  filter(year>=2010) %>%
  tidyr::pivot_wider(names_from = year, values_from = logRR13)

pheatmap(
  df_wide13[,-1],
  color = viridis(100),
  cluster_rows = TRUE,
  cluster_cols = F,
  scale = "none",
  border_color = NA,
  labels_row=df_wide13$st
)
```
ST4 trends
```{r}
s1 %>%
  filter(st %in% c('4','6B','9V','14','18C','19F','23F')) %>%
  ggplot(aes(x=year, y=log(N_IPD)))+
  geom_line()+
  facet_wrap(~st, scales ='free') +
  theme_classic()+
  geom_vline(xintercept=c(1999.5, 2009.5), lty=2, color='gray')+
  ggtitle('Trends in PCV7 serotypes')

```
```{r, fig.width=14, fig.height=12}
 readRDS('./Data/ABCs_st_1998_2021.rds') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  mutate( st= if_else(st=='16','16F', st),
          agec1 = if_else(agec %in% c("Age <2","Age 2-4") ,1,2 )) %>%
  group_by(st,agec, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup()%>%
   filter(st %in% c('4','6B','9V','14','18C','19A','19F','23F')) %>%
  ggplot(aes(x=year, y=log(N_IPD+1)))+
  geom_line()+
  facet_grid(agec~st, scales ='free') +
  theme_classic()+
  geom_vline(xintercept=c(1999.5, 2009.5), lty=2, color='gray')+
  ggtitle('Trends in PCV7 serotypes')
```


```{r}
#a1 <- read.csv('https://data.cdc.gov/api/views/en3s-hzsr/rows.csv?accessType=DOWNLOAD')

#saveRDS(a1,'./Data/ABCs_data_summary_2024_01_01.rds')

a1 <-readRDS('./Data/ABCs_data_summary_2024_01_01.rds')
```

Based on ABCs Bact-Facts, projected 3250 IPD deaths nationally in 2019. In 2019 there were 810 deaths nationally recorded as J13 (pneumococcal pneumonia) or A40.3 (pneumococcal septicemia) or G00.1 (pneumococcal meningitis) from CDC WONDER (B95.3 was not on WONDER). With the line list data we get 923 with J13, A40.3, G00.1, B95.3. There were 537 Streptococcal sepsis (A40.8/A40.9)


ABCS genomic data are in: https://www.ncbi.nlm.nih.gov/bioproject/284954

```{r}
p1 <- a1 %>%
  filter(Topic =='Case Rates' & ViewBy=='Age' & Year>=1998) %>%
 ggplot( aes(x=Year, y=Value, col=ViewBy2)) +
  geom_line() +
  theme_classic() 

ggplotly(p1)
```

```{r}
p2 <- a1 %>%
  filter(Topic =='Death Rates' & ViewBy=='Age' & Year>=1998) %>%
 ggplot( aes(x=Year, y=Value, col=ViewBy2)) +
  geom_line() +
  theme_classic() 

ggplotly(p2)
```


```{r, fig.width=12, fig.height=4}
b2019 <- read.csv('./Data/ABCs/jiac058_suppl_supplementary_table_s2.csv') %>%
  group_by(State, sero, LT5) %>%
  summarize(N_cases=n()) %>%
  mutate(sero=as.factor(sero)) %>%
  ungroup() %>%
  group_by(State, LT5) %>%
  mutate(pct = N_cases/sum(N_cases)) %>%
  ungroup()
  
ggplotly(ggplot(b2019[b2019$LT5=='neg',], aes(fill=State, y=pct, x=sero)) + 
    geom_bar(position="dodge", stat="identity") 
) 
```
correlations
```{r}

adult.c <- b2019 %>%
  filter(LT5=='neg') %>%
  ungroup() %>%
  group_by(sero) %>%
  mutate(tot_N_cases=sum(N_cases)) %>%
  ungroup() %>%
  filter(tot_N_cases >= 10) %>%
 dcast(  State ~ sero, value.var='pct') %>%
  replace(is.na(.), 0)

cor(adult.c[,-1])
```



```{r}


  p3 <- ggplot(b2019[b2019$LT5=='neg',], aes(fill = sero, values = N_cases)) +
    geom_waffle(color='white', size = 0.25) +
  facet_wrap(~State) +
  theme_classic()
ggplotly(p3)
```


```{r}
ecdc <- process_ecdc() %>%
  filter(agegrp==5 & country != 'EU/EEA' & total_cases >100 & !(st %in% c('11','11F','19','36','39','43','7','17','Other')))



ecdc.c <-  dcast(ecdc  ,country ~ st, value.var='pct') %>%
  replace(is.na(.), 0)

ecdc.c.scale <- apply(ecdc.c[,-1],2,scale)
corr.ecdc <- cor(ecdc.c.scale)

dist_mat <- dist(t(ecdc.c.scale), method = 'euclidean')
hclust_avg <- hclust(dist_mat, method = 'average')
plot(hclust_avg)
View(corr.ecdc)
```

Principal Components
```{r}

a1 <- prcomp(ecdc.c[,-1], scale = TRUE)

#reverse the signs
a1$rotation <- -1*a1$rotation

#display principal components
a1$rotation

biplot(a1, scale = 0, cex=0.5)

#calculate total variance explained by each principal component
round(a1$sdev^2 / sum(a1$sdev^2)*100)

```


## US ABCs serotype data 2016-21
https://data.cdc.gov/Public-Health-Surveillance/2016-2021-Serotype-Data-for-Invasive-Pneumococcal-/qvzb-qs6p/data_preview
```{r}
st1 <- read.csv('C:/Users/DMW63/Desktop/R_Projects/ABCs_pneumococcal_data/Data/2016-2021_Serotype_Data_for_Invasive_Pneumococcal_Disease_Cases_by_Age_Group_from_Active_Bacterial_Core_surveillance_20240101.csv') 

```

```{r}
ecdc2 <- process_ecdc2() %>%
  filter( country != 'EU/EEA' & total_cases >100 & !(st %in% c('11','11F','19','36','39','43','7','17','Other')))

write.csv(ecdc2, './Data/ecdc_processed.csv')

uk_data <- ecdc2 %>%
  filter(country=='United Kingdom'  & year %in% c(2011,2019) & st !='MISS') %>%
  group_by(agegrp,st,year) %>%
  summarize(N_IPD=sum(N_cases)) %>%
  ungroup() %>%
  group_by(agegrp,year) %>%
  arrange(-N_IPD) %>%
  mutate(stN=row_number(),
        pcv20 = if_else(st %in% c('1', '3', '4', '5', '6A', '6B', '7F', '8', '9V', '10A', '11A', '12F', '14', '15B', '18C', '19A', '19F', '22F', '23F','33F'),'PCV20','Other')) %>%
  ungroup()

```

```{r, fig.width=5, fig.height=3}
 p1 <- uk_data %>% 
   filter(agegrp==1 & year==2011) %>%
   ggplot(aes(x=stN, y=N_IPD, color=pcv20)) +
  #geom_line() +
  geom_text(aes(label=st), size=2, angle=45)+
  theme_minimal()+
  ylab('Number of isolates')+
  xlab('Serotype order')+
  ggtitle('IPD, children <5 in the UK, 2011')
p1

 p2 <- uk_data %>% 
   filter(agegrp==1 & year==2019) %>%
   ggplot(aes(x=stN, y=N_IPD, color=pcv20)) +
  #geom_line() +
  geom_text(aes(label=st), size=2, angle=45)+
  theme_minimal()+
  ylab('Number of isolates')+
  xlab('Serotype order')+
  ggtitle('IPD, children <5 in the UK, 2019')
p2

```

Australia
```{r}
a1 <- readxl::read_excel('./Data/Australia/national-notifiable-diseases-surveillance-system-nndss-public-dataset-pneumococcal-disease-invasive.xlsx', skip=1) 

a2 <- a1 %>%
  rename(agec= 'Age group', st=Serotype, year=Year, indig='Indigenous status') %>%
  group_by( indig,st, year) %>%
  summarize(N_ipd=n())

st4 <- a2 %>% filter(st=='4')

a2 %>%
  filter(st %in% c('4','19F','19A','7F','6B')) %>%
ggplot( aes(x=year, y=N_ipd)) +
  geom_line()+
  facet_wrap(indig~st, scales='free_y', nrow=3)+
  ylim(0, NA)
```

