---
title: "ABCs data import"
author: "Dan Weinberger"
date: '2022-09-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(viridis)
library(plotly)
library(scales)
library(pheatmap)
#install.packages("waffle", repos = "https://cinc.rud.is") 
library(waffle)
library(reshape2)
library(gridExtra)
source('./R/process_ecdc.R')
```



##ABCs serotype specific IPD data by age!

```{r, fig.width=5, fig.height=10}
# s1 <- read.csv('https://data.cdc.gov/api/views/qvzb-qs6p/rows.csv?accessType=DOWNLOAD')
# saveRDS(s1,'./Data/ABCs_st_1998_2021.rds') 

s1 <- readRDS('./Data/ABCs_st_1998_2021.rds') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  mutate( st= if_else(st=='16','16F', st)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup()


p1 <- s1 %>%
    group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ggplot(aes(x=year, y=log(N_IPD+0.5), group=st, color=st)) +
  geom_line()+
  theme_minimal()
ggplotly(p1)

s2_pre<- s1 %>%
  filter(year %in% c(1998, 1999)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup() %>%
    group_by(st) %>%
  summarize(N_IPD_pre=mean(N_IPD)) %>%
  ungroup() %>%
    tidyr::complete(st,  fill=list(N_IPD_pre=0))  #fills 0

s2_pre13<- s1 %>%
  filter(year %in% c(2008, 2009)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup() %>%
    group_by(st) %>%
  summarize(N_IPD_pre=mean(N_IPD)) %>%
  ungroup() %>%
    tidyr::complete(st,  fill=list(N_IPD_pre=0))  #fills 0

s2<- s1 %>%
    group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>% #sum across age group
  ungroup() %>%
    tidyr::complete(st, year, fill=list(N_IPD=0))%>%  #fills 0
  left_join(s2_pre, by='st') %>%
  mutate(N_IPD_pre = if_else(is.na(N_IPD_pre),0, N_IPD_pre) ,
    logRR = log((N_IPD+1)/(N_IPD_pre+1) )) 

max_RR <- s2 %>%
  group_by(st) %>%
  summarize(max_RR = max(logRR)) %>%
  arrange((max_RR))

s2$st <- factor(s2$st, levels = max_RR$st)

ggplot(s2, aes(x = factor(year), y = st, fill = logRR)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(x = "Year", y =
         
         "serotype", fill = "N_IPD") +
  theme_minimal()


df_wide <- s2 %>%
  dplyr::select(year, st, logRR) %>%
  tidyr::pivot_wider(names_from = year, values_from = logRR)

pheatmap(
  df_wide[,-1],
  color = viridis(100),
  cluster_rows = TRUE,
  cluster_cols = F,
  scale = "none",
  border_color = NA,
  labels_row=df_wide$st
)


```

filter more rare serotypes
```{r, fig.width=5, fig.height=10}
pcv7 <- c('4','6B','9V','14','18C','19F','23F')

pcv10gsk <- c(pcv7,'1','5','7F')

pcv13 <- c(pcv7, '1','3','5','6A','7F','19A')

# s1 <- read.csv('https://data.cdc.gov/api/views/qvzb-qs6p/rows.csv?accessType=DOWNLOAD')
# saveRDS(s1,'./Data/ABCs_st_1998_2021.rds') 

s1 <- readRDS('./Data/ABCs_st_1998_2021.rds') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  mutate( st= if_else(st=='16','16F', st)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup() %>%
  group_by(st) %>%
  mutate(cum_N= sum(N_IPD)) %>%
  filter(cum_N>100) %>%
  ungroup() %>%
  mutate(pcv13st = if_else(st %in% pcv13,'VT','NVT'))



s2_pre<- s1 %>%
  filter(year %in% c(1998, 1999)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup() %>%
    group_by(st) %>%
  summarize(N_IPD_pre=mean(N_IPD)) %>%
  ungroup() %>%
    tidyr::complete(st,  fill=list(N_IPD_pre=0))  #fills 0

s2_pre13<- s1 %>%
  filter(year %in% c(2008, 2009)) %>%
  group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup() %>%
    group_by(st) %>%
  summarize(N_IPD_pre=mean(N_IPD)) %>%
  ungroup() %>%
    tidyr::complete(st,  fill=list(N_IPD_pre=0))  #fills 0

s2<- s1 %>%
    group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>% #sum across age group
  ungroup() %>%
    tidyr::complete(st, year, fill=list(N_IPD=0))%>%  #fills 0
  left_join(s2_pre, by='st') %>%
  mutate(N_IPD_pre = if_else(is.na(N_IPD_pre),0, N_IPD_pre) ,
    logRR = log((N_IPD+1)/(N_IPD_pre+1) )) 

max_RR <- s2 %>%
  group_by(st) %>%
  summarize(max_RR = max(logRR)) %>%
  arrange((max_RR))

s2$st <- factor(s2$st, levels = max_RR$st)

df_wide <- s2 %>%
  dplyr::select(year, st, logRR) %>%
  tidyr::pivot_wider(names_from = year, values_from = logRR)

p1 <- pheatmap(
  df_wide[,-1],
  color = viridis(100),
  cluster_rows = TRUE,
  cluster_cols = F,
  scale = "none",
  border_color = NA,
  labels_row=df_wide$st
) 
p1


df_vt <- df_wide %>%
  mutate(pcv13st = if_else(st %in% pcv13,'VT','NVT')) %>%
  filter(pcv13st =='VT') %>%
    dplyr::select(-pcv13st) 

  p1 <-  pheatmap(
  df_vt[,-1] ,
  color = viridis(100),
  cluster_rows = TRUE,
  cluster_cols = F,
  scale = "none",
  border_color = NA,
  labels_row=df_vt$st,
  legend_breaks=c(-6,-4,-2,0,2,4),
  breaks=seq(from=-6, to=3.2, length.out=100)
) 

df_nvt <- df_wide %>%
  mutate(pcv13st = if_else(st %in% pcv13,'VT','NVT')) %>%
  filter(pcv13st =='NVT') %>%
    dplyr::select(-pcv13st) 

 p2 <-    pheatmap(
  df_nvt[,-1] ,
  color = viridis(100),
  cluster_rows = TRUE,
  cluster_cols = F,
  scale = "none",
  border_color = NA,
  labels_row=df_nvt$st,
  legend_breaks=c(-6,-4,-2,0,2,4),
  breaks=seq(from=-6, to=3.2, length.out=100)
) 


 rr_nvt <- s2 %>%
     mutate(pcv13st = if_else(st %in% pcv13,'VT','NVT')) %>%
     filter(pcv13st =='NVT') %>%
 ggplot( aes(x = factor(year), y = st, fill = logRR)) +
  geom_tile() +
     scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(x = "Year", y =
         "serotype", fill = "log RR") +
  theme_minimal()+
       theme(axis.text.x=element_text(angle=45, hjust=1))

 
  rr_vt <- s2 %>%
     mutate(pcv13st = if_else(st %in% pcv13,'VT','NVT')) %>%
     filter(pcv13st =='VT') %>%
 ggplot( aes(x = factor(year), y = st, fill = logRR)) +
  geom_tile() +
#     guides(fill="none")+

  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(x = "Year", y =
         "serotype", fill = "log RR") +
  theme_minimal()+
    theme(axis.text.x=element_text(angle=45, hjust=1))
```



```{r, fig.width=12, fig.height=4}
grid.arrange(grobs=list(p1[[4]], p2[[4]]), nrow = 1, top='US IPD 1998-2022')    

grid.arrange(grobs=list(rr_vt, rr_nvt), nrow = 1, top='Change in US IPD 1998-2022')    

```

## tabulate coverage of different PCVs
```{r}

pcv7 <- c('4','6B','9V','14','18C','19F','23F')

pcv10gsk <- c(pcv7,'1','5','7F')

pcv13 <- c(pcv7, '1','3','5','6A','7F','19A')

pcv15 <- c(pcv13, '22F','33F')

pcv20 <- c(pcv15, '8','10A','11A','12F','15B')

vax24 <- c(pcv20, '2','9N','17F','20') #same as MAPS24 serotypes

vax31 <- c(vax24,'15A', '16F','35B','23A','23B','31','7C')

sp0202 <- c(pcv20, '9N')

pcv25 <- c(pcv15,'8','9N','10A','12F','15BC','6C','15A','16F','24F','35B') 

v116 <- c('3','7F','19A','8','9N','10A','12F','15BC','22F','33F','15A','16F','24F','35B','6A','11A','
            17F','20','23A','23B','31')

pneumosil <- c('1','5','6A','6B','7F','9V','14','19A','19F','23F')

tab1 <- read.csv('https://data.cdc.gov/api/views/qvzb-qs6p/rows.csv?accessType=DOWNLOAD') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  mutate( st= if_else(st=='16','16F', st),
          st= if_else(st %in% c('15B','15C'),'15BC', st),
          
          flag_pcv7 = if_else(st %in% pcv7,1,0),
          flag_pcv10gsk = if_else(st %in% pcv10gsk,1,0),
          flag_pcv13 = if_else(st %in% pcv13,1,0),
          flag_pcv15 = if_else(st %in% pcv15,1,0),
          flag_pcv20 = if_else(st %in% pcv20,1,0),
          flag_vax24 = if_else(st %in% vax24,1,0),
          flag_vax31 = if_else(st %in% vax31,1,0),
          flag_sp0202 = if_else(st %in% sp0202,1,0),
          flag_pcv25 = if_else(st %in% pcv25,1,0),
          flag_v116 = if_else(st %in% v116,1,0),
          flag_pneumosil = if_else(st %in% pneumosil,1,0),
 
          ) %>%
  filter(st != 'MISS' & year==2022)

coverage_u5 <- tab1 %>%
  filter(agec %in% c('Age <2', 'Age 2-4')) %>%
  mutate(tot_IPD=sum(N_IPD)) %>%
  summarize( pcv7_N = sum(flag_pcv7*N_IPD/tot_IPD),
             pcv10_N = sum(flag_pcv10gsk*N_IPD/tot_IPD),
             pcv13_N = sum(flag_pcv13*N_IPD/tot_IPD),
             pcv15_N = sum(flag_pcv15*N_IPD/tot_IPD),
             pcv20_N = sum(flag_pcv20*N_IPD/tot_IPD),
             vax24_N = sum(flag_vax24*N_IPD/tot_IPD),
             vax31_N = sum(flag_vax31*N_IPD/tot_IPD),
             sp0202_N = sum(flag_sp0202*N_IPD/tot_IPD),
             pcv25_N = sum(flag_pcv25*N_IPD/tot_IPD),
             v116_N = sum(flag_v116*N_IPD/tot_IPD),
             peumosil_N = sum(flag_pneumosil*N_IPD/tot_IPD))
round(coverage_u5*100,1)


coverage_50plus <- tab1 %>%
  filter(agec %in% c('Age 50-64', 'Age 65+')) %>%
  mutate(tot_IPD=sum(N_IPD)) %>%
  summarize( pcv7_N = sum(flag_pcv7*N_IPD/tot_IPD),
             pcv10_N = sum(flag_pcv10gsk*N_IPD/tot_IPD),
             pcv13_N = sum(flag_pcv13*N_IPD/tot_IPD),
             pcv15_N = sum(flag_pcv15*N_IPD/tot_IPD),
             pcv20_N = sum(flag_pcv20*N_IPD/tot_IPD),
             vax24_N = sum(flag_vax24*N_IPD/tot_IPD),
             vax31_N = sum(flag_vax31*N_IPD/tot_IPD),
             sp0202_N = sum(flag_sp0202*N_IPD/tot_IPD),
             pcv25_N = sum(flag_pcv25*N_IPD/tot_IPD),
             v116_N = sum(flag_v116*N_IPD/tot_IPD),
             peumosil_N = sum(flag_pneumosil*N_IPD/tot_IPD))
round(coverage_50plus*100,1)
  

st_freq <- read.csv('https://data.cdc.gov/api/views/qvzb-qs6p/rows.csv?accessType=DOWNLOAD') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
    #filter(agec %in% c('Age 50-64', 'Age 65+')) %>%
 group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD))
  
```

```{r}
prop1 <- s1 %>%
  group_by(year) %>%
  mutate(prop_IPD=N_IPD/sum(N_IPD))
```

pre-PCV13 heatmap only
```{r, fig.width=5, fig.height=10}
df_wide <- s2 %>%
  filter(year<=2009) %>%
  dplyr::select(year, st, logRR) %>%
  tidyr::pivot_wider(names_from = year, values_from = logRR)

pheatmap(
  df_wide[,-1],
  color = viridis(100),
  cluster_rows = TRUE,
  cluster_cols = F,
  scale = "none",
  border_color = NA,
  labels_row=df_wide$st
)
```

##post PV13 only

```{r, fig.width=5, fig.height=10}
s2_13<- s1 %>%
    group_by(st, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>% #sum across age group
  ungroup() %>%
    tidyr::complete(st, year, fill=list(N_IPD=0))%>%  #fills 0
  left_join(s2_pre13, by='st') %>%
  mutate(N_IPD_pre = if_else(is.na(N_IPD_pre),0, N_IPD_pre) ,
    logRR13 = log((N_IPD+1)/(N_IPD_pre+1) )) 


df_wide13 <- s2_13 %>%
  dplyr::select(year, st, logRR13) %>%
  filter(year>=2010) %>%
  tidyr::pivot_wider(names_from = year, values_from = logRR13)

pheatmap(
  df_wide13[,-1],
  color = viridis(100),
  cluster_rows = TRUE,
  cluster_cols = F,
  scale = "none",
  border_color = NA,
  labels_row=df_wide13$st
)
```

ST4 trends
```{r}
s1 %>%
  filter(st %in% c('4','6B','9V','14','18C','19F','23F')) %>%
  ggplot(aes(x=year, y=log(N_IPD)))+
  geom_line()+
  facet_wrap(~st, scales ='free') +
  theme_classic()+
  geom_vline(xintercept=c(1999.5, 2009.5), lty=2, color='gray')+
  ggtitle('Trends in PCV7 serotypes')

```
```{r, fig.width=14, fig.height=12}
 readRDS('./Data/ABCs_st_1998_2021.rds') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  mutate( st= if_else(st=='16','16F', st),
          agec1 = if_else(agec %in% c("Age <2","Age 2-4") ,1,2 )) %>%
  group_by(st,agec, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup()%>%
   filter(st %in% c('4','6B','9V','14','18C','19A','19F','23F')) %>%
  ggplot(aes(x=year, y=log(N_IPD+1)))+
  geom_line()+
  facet_grid(agec~st, scales ='free') +
  theme_classic()+
  geom_vline(xintercept=c(1999.5, 2009.5), lty=2, color='gray')+
  ggtitle('Trends in PCV7 serotypes')
```

```{r, fig.width=6, fig.height=4}
 readRDS('./Data/ABCs_st_1998_2021.rds') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  mutate( st= if_else(st=='16','16F', st),
          agec1 = if_else(agec %in% c("Age <2","Age 2-4") ,1,2 ),
                    agec=gsub('Age ', '', agec),
          agec2 = if_else( agec %in% c('<2','2-4'), '<5',
                  if_else( agec %in% c('5-17','18-49'), '5-49',
                   if_else( agec %in% c('50-64','65+'), '50+',NA))),
                    agec2 = factor(agec2, levels=c('<5','5-49','50+'), labels=c('<5 years', '5-49 years', '50+ years') )
          ) %>%
  group_by(st,agec2, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  ungroup()%>%
   filter(st %in% c('19A','19F')) %>%
  ggplot(aes(x=year, y=N_IPD))+
  geom_line()+
  facet_wrap(st~agec2, scales ='free', nrow=2) +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1.0, hjust=1)) +
  geom_vline(xintercept=c(1999.5, 2009.5), lty=2, color='gray')+
  ggtitle('Figure 2. Trends in IPD caused by serotype 19F in the US')
```



```{r}
#a1 <- read.csv('https://data.cdc.gov/api/views/en3s-hzsr/rows.csv?accessType=DOWNLOAD')

#saveRDS(a1,'./Data/ABCs_data_summary_2024_01_01.rds')

a1 <-readRDS('./Data/ABCs_data_summary_2024_01_01.rds')
```

infer population size
```{r}

b1 <- a1 %>% 
  filter(ViewBy=='Age' & ViewBy2!='Overall' & Topic=='Case Rates') %>%
  mutate(ViewBy2=gsub('years old','', ViewBy2),
         ViewBy2=gsub('year old','', ViewBy2),
         agec = if_else(ViewBy2 %in% c('<1 ','1 '),'<2',
                if_else(ViewBy2 %in% c('5-17 '),'5-17', 
                if_else(ViewBy2 %in% c('18-34 ','35-49 '),'18-49',
                if_else(ViewBy2 %in% c('50-64 '),'50-64',  
                if_else(ViewBy2 %in% c('>=65 '),'65+', ViewBy2 
                        )
         ))))
         )

N_age <- s1 <- readRDS('./Data/ABCs_st_1998_2021.rds') %>%
  rename(agec = "Age.Group..years.",
         year=Year,
         st=IPD.Serotype,
         N_IPD = Frequency.Count) %>%
  group_by(agec, year) %>%
  summarize(N_IPD=sum(N_IPD)) %>%
  mutate(agec=gsub('Age ', '', agec)) %>%
  ungroup()
```


Based on ABCs Bact-Facts, projected 3250 IPD deaths nationally in 2019. In 2019 there were 810 deaths nationally recorded as J13 (pneumococcal pneumonia) or A40.3 (pneumococcal septicemia) or G00.1 (pneumococcal meningitis) from CDC WONDER (B95.3 was not on WONDER). With the line list data we get 923 with J13, A40.3, G00.1, B95.3. There were 537 Streptococcal sepsis (A40.8/A40.9)


ABCS genomic data are in: https://www.ncbi.nlm.nih.gov/bioproject/284954

```{r}
p1 <- a1 %>%
  filter(Topic =='Case Rates' & ViewBy=='Age' & Year>=1998) %>%
 ggplot( aes(x=Year, y=Value, col=ViewBy2)) +
  geom_line() +
  theme_classic() 

ggplotly(p1)
```

```{r}
p2 <- a1 %>%
  filter(Topic =='Death Rates' & ViewBy=='Age' & Year>=1998) %>%
 ggplot( aes(x=Year, y=Value, col=ViewBy2)) +
  geom_line() +
  theme_classic() 

ggplotly(p2)
```


```{r, fig.width=12, fig.height=4}
b2019 <- read.csv('./Data/ABCs/jiac058_suppl_supplementary_table_s2.csv') %>%
  group_by(State, sero, LT5) %>%
  summarize(N_cases=n()) %>%
  mutate(sero=as.factor(sero)) %>%
  ungroup() %>%
  group_by(State, LT5) %>%
  mutate(pct = N_cases/sum(N_cases)) %>%
  ungroup()
  
  b2019 %>% 
    filter(sero=='19F' &LT5=='neg' ) %>%
    ggplot( aes(fill=State, y=pct, x=sero)) + 
    geom_bar(position="dodge", stat="identity") +
    theme_minimal()
  
    b2019 %>% 
    filter(sero=='19F' ) %>%
    ggplot( aes(fill=State, y=pct, x=sero)) + 
    geom_bar(position="dodge", stat="identity") +
      facet_wrap(~LT5)+
    theme_minimal()
    
      b2019 %>% 
    filter(sero=='19A' ) %>%
    ggplot( aes(fill=State, y=pct, x=sero)) + 
    geom_bar(position="dodge", stat="identity") +
      facet_wrap(~LT5)+
    theme_minimal()
```

```{r}
       b2019 %>% 
    filter( LT5=='neg' & (sero=='19A' |sero=='19F')) %>%
         reshape2::dcast(.,State~sero, value.var='pct') %>%
         rename(st19a='19A',
                st19f='19F') %>%
         ggplot(aes(x=st19a, y=st19f, label=State))+
         geom_point(color='white')+
         theme_classic()+
         geom_text()
         
 
```

correlations
```{r}

adult.c <- b2019 %>%
  filter(LT5=='neg') %>%
  ungroup() %>%
  group_by(sero) %>%
  mutate(tot_N_cases=sum(N_cases)) %>%
  ungroup() %>%
  filter(tot_N_cases >= 10) %>%
 dcast(  State ~ sero, value.var='pct') %>%
  replace(is.na(.), 0)

cor(adult.c[,-1])
```



```{r}


  p3 <- ggplot(b2019[b2019$LT5=='neg',], aes(fill = sero, values = N_cases)) +
    geom_waffle(color='white', size = 0.25) +
  facet_wrap(~State) +
  theme_classic()
ggplotly(p3)
```


```{r}
ecdc <- process_ecdc() %>%
  filter(agegrp==5 & country != 'EU/EEA' & total_cases >100 & !(st %in% c('11','11F','19','36','39','43','7','17','Other')))



ecdc.c <-  dcast(ecdc  ,country ~ st, value.var='pct') %>%
  replace(is.na(.), 0)

ecdc.c.scale <- apply(ecdc.c[,-1],2,scale)
corr.ecdc <- cor(ecdc.c.scale)

dist_mat <- dist(t(ecdc.c.scale), method = 'euclidean')
hclust_avg <- hclust(dist_mat, method = 'average')
plot(hclust_avg)
View(corr.ecdc)
```

Principal Components
```{r}

a1 <- prcomp(ecdc.c[,-1], scale = TRUE)

#reverse the signs
a1$rotation <- -1*a1$rotation

#display principal components
a1$rotation

biplot(a1, scale = 0, cex=0.5)

#calculate total variance explained by each principal component
round(a1$sdev^2 / sum(a1$sdev^2)*100)

```


## US ABCs serotype data 2016-21
https://data.cdc.gov/Public-Health-Surveillance/2016-2021-Serotype-Data-for-Invasive-Pneumococcal-/qvzb-qs6p/data_preview
```{r}
st1 <- read.csv('C:/Users/DMW63/Desktop/R_Projects/ABCs_pneumococcal_data/Data/2016-2021_Serotype_Data_for_Invasive_Pneumococcal_Disease_Cases_by_Age_Group_from_Active_Bacterial_Core_surveillance_20240101.csv') 

```

```{r}
ecdc2 <- process_ecdc2() %>%
  filter( country != 'EU/EEA' & total_cases >100 & !(st %in% c('11','11F','19','36','39','43','7','17','Other')))

write.csv(ecdc2, './Data/ecdc_processed.csv')

uk_data <- ecdc2 %>%
  filter(country=='United Kingdom'  & year %in% c(2011,2019) & st !='MISS') %>%
  group_by(agegrp,st,year) %>%
  summarize(N_IPD=sum(N_cases)) %>%
  ungroup() %>%
  group_by(agegrp,year) %>%
  arrange(-N_IPD) %>%
  mutate(stN=row_number(),
        pcv20 = if_else(st %in% c('1', '3', '4', '5', '6A', '6B', '7F', '8', '9V', '10A', '11A', '12F', '14', '15B', '18C', '19A', '19F', '22F', '23F','33F'),'PCV20','Other')) %>%
  ungroup()

```

```{r, fig.width=5, fig.height=3}
 p1 <- uk_data %>% 
   filter(agegrp==1 & year==2011) %>%
   ggplot(aes(x=stN, y=N_IPD, color=pcv20)) +
  #geom_line() +
  geom_text(aes(label=st), size=2, angle=45)+
  theme_minimal()+
  ylab('Number of isolates')+
  xlab('Serotype order')+
  ggtitle('IPD, children <5 in the UK, 2011')
p1

 p2 <- uk_data %>% 
   filter(agegrp==1 & year==2019) %>%
   ggplot(aes(x=stN, y=N_IPD, color=pcv20)) +
  #geom_line() +
  geom_text(aes(label=st), size=2, angle=45)+
  theme_minimal()+
  ylab('Number of isolates')+
  xlab('Serotype order')+
  ggtitle('IPD, children <5 in the UK, 2019')
p2

```

Australia
"In mid-2018, the Australian childhood 13-valent pneumococcal conjugate vaccine schedule changed from 3+0 to 2+1, moving the third dose to 12 months of age,"

https://www.health.gov.au/resources/publications/national-notifiable-diseases-surveillance-system-nndss-public-dataset-pneumococcal-disease-invasive?language=en
```{r}
a1 <- readxl::read_excel('./Data/Australia/national-notifiable-diseases-surveillance-system-nndss-public-dataset-pneumococcal-disease-invasive.xlsx', skip=1)  %>%
  rename(agec= 'Age group', st=Serotype, year=Year, indig='Indigenous status', dx='Clinical category') %>%
      mutate( agec2 = if_else( agec %in% c('< 2 yo') ,0,
                        if_else( agec %in% c( '02-04') ,1,
                       if_else( agec %in% c('05-09', '10-14') ,2,        
                        if_else( agec %in% c('20-24', '25-29','30-34',"35-39") ,3,        
                        if_else( agec %in% c('40-44', '45-49','50-54',"55-59","60-64") ,4,        
                        if_else( agec %in% c('65-69', '70-74','75-79',"80-84","85+") ,5,999        
                               )))))),
              agec2 =factor(agec2,levels=c(0,1,2,3,4,5,999),c('<2y', '2-4y', '5-14y','20-39y','40-64y','65+y','NA'))
      )
      

 a1 %>%
  group_by( indig,st, year) %>%
  summarize(N_ipd=n()) %>%
  filter(st %in% c('4','19F','19A','7F','6B')) %>%
ggplot( aes(x=year, y=N_ipd)) +
  geom_line()+
  facet_wrap(indig~st, scales='free_y', nrow=3)+
  ylim(0, NA) +
   theme_classic()
 
  a1 %>%
  group_by( Sex,st, year) %>%
  summarize(N_ipd=n()) %>%
  filter(st %in% c('4','19F','19A','7F','6B') & Sex %in% c('Female','Male')) %>%
ggplot( aes(x=year, y=N_ipd)) +
  geom_line()+
  facet_wrap(Sex~st, scales='free_y', nrow=2)+
  ylim(0, NA)
  
  
    a1 %>%
  group_by( agec2,st, year) %>%
  summarize(N_ipd=n()) %>%
      ungroup() %>%
      group_by(agec2, st) %>%
      mutate(IPD_scaled=N_ipd/max(N_ipd)) %>%
  filter(st %in% c('4','19F','19A','7F','6B') & !is.na(agec2) ) %>%
ggplot( aes(x=year, y=IPD_scaled)) +
  geom_line()+
  facet_grid(agec2~st, scales='free_y')+
  ylim(0, NA)+
   theme_classic()
    

    
        a1 %>%
  group_by( agec2,st, year) %>%
  summarize(N_ipd=n()) %>%
      ungroup() %>%
      group_by(agec2, st) %>%
      mutate(IPD_scaled=N_ipd/max(N_ipd)) %>%
  filter(st %in% c('19F','19A') & !is.na(agec2) ) %>%
ggplot( aes(x=year, y=IPD_scaled)) +
  geom_line()+
  facet_grid(st~agec2, scales='free_y')+
  ylim(0, NA)+
   theme_classic()
    

    
 st19F_kids <-    a1 %>%
        filter(agec2==1 & st=='19F')
 
 
 vax_cov <- a1 %>%
   filter(agec=="< 2 yo") %>%
   rename(pcv7="Prevenar (7vPCV)" ,pcv10="Synflorix (10vPCV)", pcv13="Prevenar 13 (13vPCV)") %>%
   group_by(year) %>%
   mutate(pcv7 =as.numeric(pcv7),
          pcv13=as.numeric(pcv13),
          pcv10=as.numeric(pcv10),
          pcv7_receipt=(pcv7>0),
          pcv13_receipt=(pcv13>0),
          pcv10_receipt=(pcv10>0))%>%
   summarize(pcv7=mean(pcv7_receipt, na.rm=T), pcv13=mean(pcv13_receipt, na.rm=T), pcv10=mean(pcv10_receipt, na.rm=T))
```

```{r, fig.width=5, fig.height=2}
        a1 %>%
  group_by( agec2,st, year) %>%
  summarize(N_ipd=n()) %>%
      ungroup() %>%
      group_by(agec2, indig,st) %>%
      mutate(IPD_scaled=N_ipd/max(N_ipd)) %>%
  filter(st %in% c('19F') & agec2 !='NA' ) %>%
ggplot( aes(x=year, y=IPD_scaled)) +
  geom_line()+
  facet_grid(~agec2, scales='free_y')+
  ylim(0, NA)+
   theme_classic()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
    
```


